Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-06-30 manuscript_agu_rev1.tex
Update on Overleaf.

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. However, removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

Because the analyzed ICON simulation was free-running (years 2021--2024, inclusive), weather and climate oscillations (such as the El Niño--Southern Oscillation phase) are not expected to be equivalent to reality at the same time and place. To compare with the observations collected during a different time period (years 2010--2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option (\url{https://alcf.peterkuma.net/documentation/cli/cmd_model.html}). For radiosonde profiles, the same mapping of time was done. That is, when selecting an equivalent profile from the model, the time of the profile was changed so that the time relative to the start of the year was preserved, but the year was changed to one of the four years available in the model data. Thus, for every radiosonde launch, there were four equivalent model profiles. The geographical location was kept the same.

@@

In our analysis we partitioned our dataset by cyclonic activity and stability into multiple subsets in order to evaluate cloud biases in the context of the main physical controlling processes. The SO is a region of occurrence of both extratropical and polar cyclones. Cyclonic activity results in cloud formation at the air mass boundaries along the cold and warm fronts, as well as inside the cold sector, after a passing cold sector destabilizes the atmosphere relative to the surface temperature. In the cold front and cold sector, clouds are convectively driven, including deep convection, and the advection of colder air masses over warmer ocean surfaces can trigger convection and subsequent cloud formation. Conversely, warm advection can trigger fog or cloud formation by boundary layer air cooled by the ocean surface until it reaches saturation. More quiescent areas outside of cyclones can also be associated with clouds. These can be, for example, associated with clouds formed by warm or cold advection outside of cyclones, persistent clouds, clouds formed due to diurnal heating or cooling, or due to ocean currents. Boundary layer stability can be expected to be associated with clouds by either allowing convection and turbulence under weak stability, inhibiting convection turbulence under strong stability, and by capping inversion controlling the cloud top height or trapping moist air near the surface and preventing fog dispersion. Therefore, dividing our dataset by these subsets allows us to quantify model biases associated with some of the main physical processes controlling cloud formation, persistence, and dissipation. Other methods of subsetting, such as using the ISCCP pressure--optical thickness diagram \cite{rossow1991,rossow1999,hahn2001} to separate profiles by cloud regimes and other cloud regime classifications \cite{oreopoulos2016,schuddeboom2018}, would be feasible.

@@

Fig.~\ref{fig:rs-thetav-hur-cloud} shows virtual potential temperature and RH profiles for profiles containing fog, cloud at 500 m, and cloud at 1.5 km. These situations are characterized by particular cloud biases as identified in the lidar cloud occurrence analysis. The rationale is to examine $\theta_v$ and RH associated with these situations. Foggy situations are characterized by a rapid increase of $\theta_v$ with height and an observed average RH of about 90\% near the surface (Fig.~\ref{fig:rs-thetav-hur-cloud}a). In contrast, the models simulate higher RH in the first 100~m under foggy conditions by several percentage points. In situations with clouds occurring at 500 m, $\theta_v$ is relatively flat between the surface and 500 m (Fig.~\ref{fig:rs-thetav-hur-cloud}b), as expected for convectively driven clouds. The observed RH peaks at 500 m at about 90\%. The models, however, simulated higher RH between the surface and 500 m under these conditions. ICON and ERA5 show a stronger decrease of RH above this height than observations, and ERA5 shows more strongly stable stratification. Unlike the foggy and 500-m cloud situations, situations with clouds at 1.5 km do not have a flat $\theta_v$ with height. This indicates that unlike the former, clouds at 1.5 km are not (or not as strongly) convectively driven. As expected, RH in these situations peaks at 1.5 km at about 85\% in observations. In the models, this peak is much less pronounced.

@@

Fig.~\ref{fig:rs-thetav-hist} shows histograms the same as the previous figure, but for virtual potential temperature. They show a more complex picture, characterized by a central peak at about 0°C near the surface, increasing to about 5°C at 2 km (Fig.~\ref{fig:rs-thetav-hist}a). For cloudy bins, the central peak is generally flatter with height and even shows a minimum in $\theta_v$ at about 500 m (Fig.~\ref{fig:rs-thetav-hist}i). This is indicative of convection being associated with clouds at these heights, which results in flat virtual potential temperature profiles. In the reanalyses, in the first 200 m, values slightly above 0°C are associated with more clear bins than in observations, and values slightly below 0°C with fewer (Fig.~\ref{fig:rs-thetav-hist}g--h). Conversely, the opposite is true for cloudy bins (Fig.~\ref{fig:rs-thetav-hist}k--l). Situations with 0°C near-surface air temperature might occur predominantly when open ocean surface keeps the near-surface air temperature close to 0°C under otherwise colder air mass conditions, such as under cold advection. ICON displays a notable bias above about 1 km, where the central peak is strongly underestimated (Fig.~\ref{fig:rs-thetav-hist}j). Instead, these heights and values of $\theta_v$ are more associated with clear bins (Fig.~\ref{fig:rs-thetav-hist}f). This might be related to the strong underestimation of cloud occurrence at these heights.

@@

The campaigns show remarkably similar biases in cloud occurrence by height in the lidar comparison (Fig.~\ref{fig:cloud-occurrence-panel}), which indicates that common underlying causes for the biases exist regardless of longitude and season. ICON underestimates the total cloud fraction by about 10\%, with an overestimation of clouds below 2~km and an underestimation of clouds above 2~km. The reanalyses underestimate the total cloud fraction by about 20\%. ERA5 overestimates cloud below 1~km but underestimates near-surface cloud and fog. ICON strongly overestimates the peak of cloud occurrence at about 500 m. This can be explained by the radiosonde comparison, showing that it is too moist at around this height (Fig.~\ref{fig:potential-temperature}a); has underestimated LTS (Fig.~\ref{fig:cyclone-stability} and \ref{fig:rs-lts-hist}), permitting shallow convection to this height; and has underestimated near-surface RH (Fig.~\ref{fig:stats-hist-surf}), resulting in higher LCL (Fig.~\ref{fig:cloud-occurrence}). Similar to our results for mid-level clouds, \citeA{cesana2022} showed that CMIP6 models also tend to underestimate cloud occurrence above 2~km over the SO, although their analysis in this case was limited to liquid clouds.

>>>>

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

Because the analyzed ICON simulation was free-running (years 2021--2024, inclusive), weather and climate oscillations (such as the El Niño--Southern Oscillation phase) are not expected to be equivalent to reality. To compare with the observations collected during a different time period (years 2010--2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option (\url{https://alcf.peterkuma.net/documentation/cli/cmd_model.html}). For radiosonde profiles, the same mapping of time was done. That is, when selecting an equivalent profile from the model, the time of the profile was changed so that the time relative to the start of the year was preserved, but the year was changed to one of the four years available in the model data. Thus, for every radiosonde launch, there were four equivalent model profiles. The geographical location was kept the same.

@@

In our analysis, we partitioned our dataset by cyclonic activity and stability into multiple subsets to evaluate cloud biases in the context of the main physical controlling processes. The SO is a region of the occurrence of both extratropical and polar cyclones. Cyclonic activity results in cloud formation at the air mass boundaries along the cold and warm fronts, as well as inside the cold sector, after a passing cold sector destabilizes the atmosphere relative to the surface temperature. In the cold front and cold sector, clouds are convectively driven, including deep convection, and the advection of colder air masses over warmer ocean surfaces can trigger convection and subsequent cloud formation. In contrast, warm advection can trigger fog or cloud formation by boundary layer air cooled by the ocean surface until it reaches saturation. More quiescent areas outside of cyclones can also be associated with clouds. These can be, for example, associated with clouds formed by warm or cold advection outside of cyclones, persistent clouds, clouds formed due to diurnal heating or cooling, or due to ocean currents. Boundary layer stability can be expected to be associated with clouds by either allowing convection and turbulence under weak stability, inhibiting convection turbulence under strong stability, and by capping inversion controlling the cloud top height or trapping moist air near the surface and preventing fog dispersion. Therefore, dividing our dataset by these subsets allows us to quantify model biases associated with some of the main physical processes controlling cloud formation, persistence, and dissipation. Other methods of subsetting, such as using the ISCCP pressure--optical thickness diagram \cite{rossow1991,rossow1999,hahn2001} to separate profiles by cloud regimes and other cloud regime classifications \cite{oreopoulos2016,schuddeboom2018}, would be feasible.

@@

Fig.~\ref{fig:rs-thetav-hur-cloud} shows virtual potential temperature and RH profiles for profiles containing fog, cloud at 500 m, and cloud at 1.5 km. These situations are characterized by particular cloud biases as identified in the lidar cloud occurrence analysis. The rationale is to examine $\theta_v$ and RH associated with these situations. Foggy situations are characterized by a rapid increase of $\theta_v$ with height and an observed average RH of about 90\% near the surface (Fig.~\ref{fig:rs-thetav-hur-cloud}a). In contrast, the models simulate higher RH in the first 100~m under foggy conditions by several percentage points. In situations with clouds occurring at 500 m, $\theta_v$ is relatively constant between the surface and 500 m (Fig.~\ref{fig:rs-thetav-hur-cloud}b), as expected for convectively driven clouds. The observed RH peaks at 500 m at about 90\%. The models, however, simulated higher RH between the surface and 500 m under these conditions. ICON and ERA5 show a stronger decrease of RH above this height than observations, and ERA5 shows more strongly stable stratification. Unlike the foggy and 500-m cloud situations, situations with clouds at 1.5 km do not have a flat $\theta_v$ with height. This indicates that unlike the former, clouds at 1.5 km are not (or not as strongly) convectively driven. As expected, RH in these situations peaks at 1.5 km at about 85\% in observations. In the models, this peak is much less pronounced.

@@

Fig.~\ref{fig:rs-thetav-hist} shows histograms the same as the previous figure, but for virtual potential temperature. They show a more complex picture, characterized by a central peak at about 0°C near the surface, increasing to about 5°C at 2 km (Fig.~\ref{fig:rs-thetav-hist}a). For cloudy bins, the central peak is generally more constant with height and even shows a minimum in $\theta_v$ at about 500 m (Fig.~\ref{fig:rs-thetav-hist}i). This is indicative of convection being associated with clouds at these heights, which results in flat virtual potential temperature profiles. In the reanalyses, in the first 200 m, values slightly above 0°C are associated with more clear bins than in observations, and values slightly below 0°C with fewer (Fig.~\ref{fig:rs-thetav-hist}g--h). Conversely, the opposite is true for cloudy bins (Fig.~\ref{fig:rs-thetav-hist}k--l). Situations with 0°C near-surface air temperature might occur predominantly when open ocean surface keeps the near-surface air temperature close to 0°C under otherwise colder air mass conditions, such as under cold advection. ICON displays a notable bias above about 1 km, where the central peak is strongly underestimated (Fig.~\ref{fig:rs-thetav-hist}j). Instead, these heights and values of $\theta_v$ are more associated with clear bins (Fig.~\ref{fig:rs-thetav-hist}f). This might be related to the strong underestimation of cloud occurrence at these heights.

@@

The campaigns show remarkably similar biases in cloud occurrence by height in the lidar comparison (Fig.~\ref{fig:cloud-occurrence-panel}), which indicates that common underlying causes for the biases exist regardless of longitude and season. ICON underestimates the total cloud fraction by about 10\%, with an overestimation of clouds below 1~km and an underestimation of clouds above 1~km. The reanalyses underestimate the total cloud fraction by about 20\%. ERA5 overestimates cloud below 1~km but underestimates near-surface cloud and fog. ICON strongly overestimates the peak of cloud occurrence at about 500 m. This can be explained by the radiosonde comparison, showing that it is too moist at around this height (Fig.~\ref{fig:potential-temperature}a); has underestimated LTS (Fig.~\ref{fig:cyclone-stability} and \ref{fig:rs-lts-hist}), permitting shallow convection to this height; and has underestimated near-surface RH (Fig.~\ref{fig:stats-hist-surf}), resulting in higher LCL (Fig.~\ref{fig:cloud-occurrence}). Similar to our results for mid-level clouds, \citeA{cesana2022} showed that CMIP6 models also tend to underestimate cloud occurrence above 2~km over the SO, although their analysis in this case was limited to liquid clouds.
